---
output:
  pdf_document: default
  html_document: default
---
Wine Dataset Exploration by Hajeong Noh
========================================================
This project is my first Exploratory Data Analysis using R. The dataset includes
several features of wines and its quality. The format includes Univariate, 
Bivariate and Multivariate analysis with a final summary and reflection. The 
original dataset is available [here](http://www3.dsi.uminho.pt/pcortez/wine/)

```{r echo=FALSE, message=FALSE, warning=FALSE, Packages}
library(ggplot2)
library(dplyr)
library(tidyr)
library(PerformanceAnalytics)
library(reshape2)
library(gridExtra)
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Load_the_Data}
# Load the Data
red <- read.csv('wineQualityReds.csv')
white <- read.csv('wineQualityWhites.csv')

# Add a type attribute
red$type = 'red'
white$type = 'white'

# bind rows of two data frames
df <- rbind(red, white)

# drop X column which is unnecessary for analysis
df <- df %>% dplyr::select(-X)
```

# Univariate Plots Section
Let's take a look at the dimensions, structure and summary of wine dataset.

```{r echo=FALSE, message=FALSE, warning=FALSE, Wine_Dimension}
dim(df)
```


```{r echo=FALSE, message=FALSE, warning=FALSE, Wine_Structure}
str(df)
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Wine_Summary}
summary(df)
```

The key stat quality has median value 6. minimum is 3 and maximum is 9. Even 
though quality is ranging from 0 to 10, only 3 through 9 are used.

Most density values fall between 0.99 and 1.

There is a wine that contains residual sugar way more than other wines.
According to the guideline, this wine is sweet (residual sugar is more than 45).

```{r echo=FALSE, message=FALSE, warning=FALSE, Quality_Histogram}
ggplot(aes(x=quality), data=df) +
  geom_bar() +
  scale_x_continuous(breaks=seq(3,9,1))
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Quality_Table}
table(df$quality)
```

The most common wine quality rating is 6 followed by 5 and 7. Few observations 
are at 3 and 9, which can be considered as very poor and very good, 
respectively.

```{r echo=FALSE, message=FALSE, warning=FALSE, Type_Histogram}
ggplot(aes(x=type), data=df) +
  geom_bar()
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Type_Table}
table(df$type)
```

White wine is approximately 3 times more observed in the dataset.

```{r echo=FALSE, message=FALSE, warning=FALSE, Variable_Distribution_Each}
ggplot(df %>% dplyr::select(-12:-13) %>% gather(), aes(value)) +
  geom_histogram() +
  facet_wrap(~key, scales='free')
```

The histograms are relatively located on the left side of each plot, since there
are outliers with high values. Let's exclude outliers to observe distributions 
better.

```{r echo=FALSE, message=FALSE, warning=FALSE, Percentage_Quantile_for_Each}
# Make quantile table
df.quantile <- df %>% dplyr::select(-12:-13) %>% gather() %>% group_by(key) %>% 
  summarise(upper = quantile(value, .99))

# Merge quantile table and data for histogram
mergeForHist <- merge(df %>% dplyr::select(-12:-13) %>% gather(), df.quantile, 
                      by='key')

# Delete rows where the data is bigger than 99% quantile
histData <- mergeForHist %>% filter(value <= upper)
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Histogram_without_Outliers}
ggplot(histData, aes(x=value)) +
  geom_histogram() +
  facet_wrap(~key, scales='free')
```

Most of the variables follow normal distribution except residual sugar. Residual
sugar distribution follow long-tail distribution. Let's transform the x-axis to 
observe the distribution better.

```{r echo=FALSE, message=FALSE, warning=FALSE, Sugar_Histogram_Transformed}
ggplot(aes(x=residual.sugar), 
       data=subset(df, residual.sugar < quantile(df$residual.sugar, 0.99))) +
  geom_histogram() +
  scale_x_log10(breaks=c(1, 3, 10)) +
  geom_vline(xintercept=3, linetype='longdash')
```

After x-axis transformation, the distribution follows bimodal. I can divide the 
wines into 2 groups of which has residual sugar less than 3 and which has more
than 3 for further analysis.

# Univariate Analysis

### What is the structure of your dataset?

There are 6,497 observations with 11 quantitative features (alcohol, chlorides,
citric acid, density, fixed acidity, free sulfur dioxide, pH, residual sugar, 
sulphates, total sulfur dioxide and volatile acidity), quality (discrete 
variable) and type (categorical).

Other observations:
- The most observed value of quality is 6 followed by 5 and 7.
- The average alcohol percentage is 10.51%.
- About 75% of white wines have residual sugar less than 9.9.
- The median citric acid is 0.32.
- The bimodality of residual histogram shows that we can group the wines with 
more sugar and less sugar.


### What is/are the main feature(s) of interest in your dataset?

The main features in the dataset is quality and type. I¡¯d like to explore which 
features show the most difference as the type differs. I also want to determine 
which feature contributes the most for predicting the quality of a wine.

### What other features in the dataset do you think affect the quality of wine?

I expect that residual sugar and citric acid contribute to the quality of white
wines. I think cholorides may also play a significant role.

### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?

I made an adjustment when I tried to see each histogram of a variable. I found 
out many histograms are located on the left side of plot, so I removed outliers
to check the distribution clearly.

Another adjustment I did was transform the x-axis of residual sugar histogram,
since the original histogram show long-tailed distribution. After transformation
I could observe that the distribution follows bimodal, which shows that I can 
divide less and more sugar group for further analysis.

# Bivariate Plots Section

```{r echo=FALSE, message=FALSE, warning=FALSE, Correlation_Matrix}
PerformanceAnalytics::chart.Correlation(df %>% dplyr::select(-type), 
                                        histogram=TRUE)
```

The matrix above is plotted using PerformanceAnalytics package. Here are some of
the correlations involving quality.

- Quality and Alcohol: 0.44
- Quality and Density: -0.31
- Quality and Volatile acidity: -0.27

None of those show a strong correlation.

Other interesting correlations are observed between other features. 

- Free sulfur dioxide and Total sulfur dioxide: 0.72
- Density and Alcohol: -0.69
- Density and Residual sugar: 0.55

Let's take a look ate the relationship between features and quality in more 
detail.

```{r echo=FALSE, message=FALSE, warning=FALSE, Alcohol_Quality_Scatter}
ggplot(aes(x=alcohol, y=quality), 
       data=subset(df, alcohol < quantile(df$alcohol, .99))) +
  geom_jitter(alpha=.1) +
  scale_y_continuous(breaks=seq(3,9,1)) +
  geom_smooth(method='lm')
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Density_Quality_Scatter}
ggplot(aes(x=density, y=quality), 
       data=subset(df, density < quantile(df$density, .99))) +
  geom_jitter(alpha=.1) +
  scale_y_continuous(breaks=seq(3,9,1)) +
  geom_smooth(method='lm')
```

The scatter plots above show the relationship of features which showed the 
moderate correlation with quality (0.7 > cor >= 0.3) in the correlation matrix. 
I jittered the plots to visualize trends better and exclude top 1% outliers to 
check the general trend.

The first scatter plot shows there is a positive relationship between alcohol 
and quality, even though the correlation is not so strong.

The second scatter plot shows on the other hand, there is a negative 
relationship between density and quality. Heavy wine can harm the taste.

What about the other variables that I expected to affect the quality?
Let's find out!

```{r echo=FALSE, message=FALSE, warning=FALSE, Sugar_Density_Scatter}
ggplot(aes(x=residual.sugar, y=quality), 
       data=subset(df, residual.sugar < quantile(df$residual.sugar, .99))) +
  geom_jitter(alpha=.1) +
  scale_y_continuous(breaks=seq(3,9,1)) +
  geom_smooth(method='lm')
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Chlorides_Density_Scatter}
ggplot(aes(x=chlorides, y=quality), 
       data=subset(df, chlorides < quantile(df$chlorides, .95))) +
  geom_jitter(alpha=.1) +
  scale_y_continuous(breaks=seq(3,9,1)) +
  geom_smooth(method='lm')
```

The correlation of residual sugar and quality is 0.07, which is very weak.
Moreover, the distribution looks very even.

The second plot shows that chlorides and quality has a negative relationship.
However, the correlation of chlorides and quality is -0.20, which shows a weak 
relationship between features.

Now let's search for an interesting scatter plot between features except 
quality. Drawing plots which showed high correlations in the correlation matrix
will be interesting.

```{r echo=FALSE, message=FALSE, warning=FALSE, Strong_Correlation_Plots}
p1 = ggplot(aes(x=free.sulfur.dioxide, y=total.sulfur.dioxide), data=df) +
  geom_point(alpha=.1) +
  xlim(0, quantile(df$free.sulfur.dioxide, 0.99)) +
  ylim(0, quantile(df$total.sulfur.dioxide, 0.99))

p2 = ggplot(aes(x=alcohol, y=density), data=df) +
  geom_point(alpha=.1) +
  xlim(8, quantile(df$alcohol, 0.99)) +
  ylim(0.987, quantile(df$density, 0.99))

p3 = ggplot(aes(x=residual.sugar, y=density), data=df) +
  geom_point(alpha=.1) +
  xlim(0, quantile(df$residual.sugar, 0.99)) +
  ylim(0.987, quantile(df$density, 0.99))

grid.arrange(p1, p2, p3, ncol=3)
```

The plots above show the distribution between features that has a relatively 
strong correlation. Free sulfur dioxide and total sulfur dioxide evidently has
a strong positive correlation 0.72.

Another interesting realationships are density with alcohol and residual sugar.
Density tends to go down as alcohol percentage gets higher and as residual sugar
is less contained in the wine. I assume that alcohol is less dense than other 
ingredients and residual sugar is more dense.

Now, let's make plots that shows the difference of red wine and white wine!

```{r echo=FALSE, message=FALSE, warning=FALSE, Density_alcohol_Correlation}
meltDataByType = df %>% reshape2::melt(id.vars=c("type"))
ggplot(aes(x=type, y=value), data=meltDataByType) +
  facet_wrap(~variable, scales='free') +
  geom_boxplot()
```

The box plots above provide very interesting intuitions about the difference of 
red wine and white wine. While there is no big difference between quality and 
alcohol in both wines, white wines includes has less acidity and but more sugar 
and sulfur dioxide in average. 

# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation.

Quality correlates moderately with alcohol and density. 

As alcohol percentage increases, the quality tends to increase as well. However,
the slope is very gentle. The correlation efficient of quality and alcohol is
0.436.

On the other hand, as density increases, the quality tends to decrease. The 
correlation efficient of quality and density is -0.307.

I also found there are a difference in ingredients as the type of wine differs.
While there is no big difference between quality and alcohol in red and white 
wines, white wines includes has less acidity and but more sugar and sulfur 
dioxide in average. 

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

I observed that density correlates strongly with alcohol and residual sugar.

As residual sugar increases, the density increase as well.The relationship 
appears to be linear. The correlation efficient of density and alcohol is 0.839.

On the other hand, as alcohol increases, density tends to decrease. The 
relationship between alcohol and density appears to be linear. The  correlation 
efficient of quality and density is -0.780.

### What was the strongest relationship you found?

The strongest relationship that was between free sulfur dioxide and total sulfur
dioxide with correlation 0.72. This seems to be a logical result.

# Multivariate Plots Section

Let's take a look at the scatter plots between quality and other features with 
colored type.

```{r echo=FALSE, message=FALSE, warning=FALSE, Scatter_Plots_With_Type}
meltDataByTypeAndQuality = df %>% reshape2::melt(id.vars=c("type", "quality"))
ggplot(aes(x=value, y=quality, color=type), data=meltDataByTypeAndQuality) +
  facet_wrap(~variable, scales='free') +
  geom_jitter(alpha=.1)
```

As we have seen in the boxplots, we can see some separation of red wines and 
white wines in each box plots. red wines are observed in more acidity but less
sulfur dioxide area. However, in the residual sugar-quality plot, it is not so 
clear how the data points are distributed. Let's draw a seprate plot.

```{r echo=FALSE, message=FALSE, warning=FALSE, Sugar_Diff_By_Wine_Type}
ggplot(aes(x=residual.sugar, y=quality, color=type), 
       data=subset(df, residual.sugar < quantile(df$residual.sugar, 0.99))) +
  geom_jitter(alpha=.1) +
  geom_vline(xintercept=3, linetype='longdash') +
  xlim(0, quantile(df$residual.sugar, 0.99)) +
  scale_x_log10(breaks=c(1,3,10)) +
  scale_y_continuous(breaks=seq(3,9,1))
```

X-axis is scaled to logarithm, as we did for residual sugar it in the univariate
section. The dashed line is where residual sugar is 3. We can observe that the 
most red wines are located on the left side of the dashed line. I think red
wines containes relatively less sugar and white wines has diverse sorts of wines
in terms of residual sugar amount.

Let's plot the most correlated features with quality, alcohol and density with
quality.

```{r echo=FALSE, message=FALSE, warning=FALSE, Quality_Distribution}
ggplot(aes(x=alcohol, y=density, color=as.factor(quality)), 
       data=subset(df, density < quantile(df$density, 0.99))) +
  geom_point() +
  scale_color_brewer()
```

As we observed in the bivariate section, alcohol and density scatter plot shows
a negative relationship. lighter blue points are observed more on the left
upper part of the plot, which represents less alcohol and high density wines get
low ratings by judges.

```{r echo=FALSE, message=FALSE, warning=FALSE, Volatile_Acidity_Diff_By_Type}
ggplot(aes(x=quality, y=volatile.acidity), data=df) +
  geom_line(aes(color=type), stat='summary', fun.y=median) +
  scale_x_continuous(breaks=seq(3,9,1))
```

Based on the correlation matrix, volatile acidity and quality has a negative 
relationship with correlation efficient -0.27. However, we can see that the 
quality change line appears totally different by the wine type.

The median volatile acidity of red wines decreases as the quality of red wine 
increases. On the other hand, the median volatile acidity of white wines does 
not show big differences as the quality changes.

I can assume that the volatile acidity is a good criteria to measure a quality 
of a red wine, but not for a white wine.

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

Using multivariate Analysis, I was able to visualize how the data points of red
wines and white wines are located. I could find out that the residual sugar of
red wines are mostly below 3. Also, the relationship between quality, alcohol 
and density is strengthened.

### Were there any interesting or surprising interactions between features?

I was able to observe that the average volatile acidity trend as the quality 
increases totally differs by the type of wine. I found out the volatile acidity 
is a good criteria to measure a quality of a red wine, but not for a white wine.

------

# Final Plots and Summary

### Plot One
```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_One}
p1 = ggplot(aes(x=residual.sugar), 
            data=subset(df, residual.sugar<quantile(df$residual.sugar, 0.99))) +
  geom_histogram() +
  xlim(0, quantile(df$residual.sugar, 0.5)) +
  scale_x_log10(breaks=c(1,3,10)) +
  geom_vline(xintercept=3, linetype='longdash') +
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank()) +
  ylab('Count of wines')

p2 = ggplot(aes(x=residual.sugar, y=quality, color=type), 
            data=subset(df, residual.sugar<quantile(df$residual.sugar, 0.99))) +
  geom_jitter(alpha=.1) +
  geom_vline(xintercept=3, linetype='longdash') +
  xlim(0, quantile(df$residual.sugar, 0.99)) +
  scale_x_log10(breaks=c(1,3,10)) +
  scale_y_continuous(breaks=seq(3,9,1)) +
  theme(legend.position = 'bottom') +
  ylab('Quality ratings (1 to 10)') +
  xlab('Residual sugar (g / dm^3)')

p2$labels$colour = 'Wine type'

grid.arrange(p1, p2, 
             top='Wine Classfication according to Residual Sugar Amount')
```

### Description One

The distribution of residual sugar of wines appear to be bimodal on log scale.
So I divided the group to the one that has residual sugar less than 3 and the 
other that has residual sugar more than 3.

By plotting it in a scatter plot with coloring the types, I discovered residual 
sugar and quality has no noticeable relationship. Also, most of the red wines 
are observed in the less residual sugar group. And white wines are observed 
quite evenly over all the residual sugar range.


### Plot Two
```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_Two}
p3 <- ggplot(aes(x=alcohol, y=density, color=as.factor(quality)), 
       data=subset(df, density < quantile(df$density, 0.99))) +
  geom_point() +
  scale_color_brewer() +
  ggtitle('Scatter Plot of Density vs Alcohol with Color Set by Quality') +
  xlab('Alcohol (%)') +
  ylab('Density (g / cm^3)')

p3$labels$colour <- 'Quality rating'

p3
```

### Description Two

The plot reflects the relationship between alcohol, density and quality in a 
single plot.

In general, density tends to decrease as alcohol percentage increases, even 
though the features are not strongly correlated. I guess it is because alcohol 
is less dense than other ingredients. Also, we can observe that light blue data 
points are more observed on the left upper part of plot, which represents that 
less alcohol and high density wines are relatively poorly rated.

### Plot Three
```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_Three}
p4 <- ggplot(aes(x=quality, y=volatile.acidity), data=df) +
  geom_line(aes(color=type), stat='summary', fun.y=median, size=2) +
  scale_x_continuous(breaks=seq(3,9,1)) +
  ggtitle('Median of Volatile Acidity for each Quality Rating by Wine Type') +
  xlab('Quality rating (1 to 10)') +
  ylab('Volatile Acidity (g / cm^3)')

p4$labels$colour = 'Wine type'

p4
```

### Description Three

The Median of volatile acidity for each quality rating differs evidently by the 
type of wine. 

The median volatile acidity of red wines decreases as the quality of red wine 
increases. On the other hand, the median volatile acidity of white wines does 
not show big differences as the quality changes.

Perhaps, the volatile acidity is a good criteria to measure a quality of a red
wine, but not for a white wine.

------

# Reflection

The wine data set contains information on 6,497 wines across 13 variables. I 
started to explore the dataset by plotting variables and continued to search for
interesting relationships. Eventually I discovered some interesting relationship
between quality, type and other features.

In early phase of exploration, I thought there must be a main variable that 
explaines the quality rating of wines. I was surprised that it is possible that 
it is hard to explain the quality of variable even if I consider every variable 
in the dataset.

Also, I discovered the general trend of a variable can be a lot different when 
you seperate another variable and observe the trends of each seprated variable.
For example, volatile acidity and quality shows a negative correlation in 
general. However, if you divide the dataset into red wine and white wine and 
observe the volatile acidity by quality, you can see red wine shows a rapid 
decrease in volatile acidity as quality rating increases. On the other hand,
the volatile acidity of white wine cannot explain much about the change of 
quality ratings.

To investigate the data further, I would like to get the information of how long
the wines are aged. Normally well-aged wines are considered as quality wines, so
I think this variable could predict the quality of wine better. Also, it would 
be interesting to trace the change of ingredients as the wine ages.